Implementation and assessment of tools for integrating forest roads into projections of the cumulative effects of disturbance on wildlife

### 2024-05-27 analysis/paper/paper.Rmd
minor edits based on Greg's comments, more to come on bigger things

institute:
  - ECCC: Landscape Science and Technology Division, National Wildlife Research Center, Environment and Climate Change Canada, Ottawa, ON, Canada  
  - FLNR: Ministry of Forest Lands, Natural Resource Operations and Rural Development, Victoria, BC, Canada
  - UBC:  Faculty of Forestry, Forest Resources Management, The University of British Columbia, Vancouver, BC, Canada

@@

We expect these simple graph-based road projection methods to represent a reasonable compromise between computational cost and realism for stochastic projections on large landscapes, and we expect our simplified method for penalizing steep grades to yield more realistic road networks in steep terrain. We do not have a similarly strong hypothesis about whether minimimum spanning tree projections are likely to be more or less realistic than interative least cost path projections. If designers of real roads take a larger view to accessing multiple sites at the lowest cost we might expect minimum spanning tree algorithms to yield more realistic networks. However, real road networks are built over time, and it is not necessarily true that they designed to minimize overall cost, so it is possible that larger more costly road networks built by iterative least-cost path algorithms will be more realistic.

@@

We chose the Revelstoke timber supply area (TSA 27) in British Columbia, Canada as an assessment area because it is mountainous (Figure \@ref(fig:study-area)), containing approximately 38% non-forested land [@british_columbia_ministry_of_forests_revelstoke_2022] where road building is difficult or impossible. Thus, the area provides opportunities to observe the behaviour of road projection algorithms in steep complex terrain. Forestry is the dominant resource extraction activity and much of the road network was built to access forest harvest cutblocks, allowing projection of the road network from historic forest harvest data. The Revelstoke TSA includes portions of the Monashee and Selkirk mountain ranges, and is bisected by the Columbia River. At low to mid-elevations, forests of the interior cedar-hemlock biogeoclimatic zone [@british_columbia_ministry_of_forests_revelstoke_2022] include remnants of rare old growth inland temperate rainforest [@government_of_canada_parks_canada_agency_old_2017]. At higher elevations, forests are dominated by Englemann spruce and subalpine fir [@british_columbia_ministry_of_forests_revelstoke_2022]. The area is within the ancestral territories of the Ktunaxa Nation, Secwepemc Nation, Syilx and Sinixt [@british_columbia_ministry_of_forests_revelstoke_2022]. Two of six southern mountain woodland caribou herds in the area have been extirpated, and the remaining herds are considered Endangered by the Committee on the Status of Endangered Wildlife in Canada and listed as threatened under the Canadian Species at Risk Act [@palm_long_2020; @british_columbia_ministry_of_forests_revelstoke_2022; @environment_canada_recovery_2014]. Mountain caribou avoid roads leading to a loss of available habitat [@polfus_identifying_2011; @apps_factors_2006], suffer increased mortality near roads [@apps_spatial_2013], and decline in abundance as the footprint of roads and cutblocks increases [@lochhead_linking_2022]. Roads also alter grizzly habitat use, home range selection, movements, population fragmentation, survival, reproductive rates, population density, trends, and conservation status [@proctor_effects_2020].

@@

We focus on methods that link multiple target locations to an existing road network by finding minimum-cost locations for new roadways that link landings within harvest cutblocks (i.e., log decks) to the existing road network, described by Dean [-@dean_finding_1997] as the multiple target access problem (MTAP). Specifically, we compare iterative least cost path (ILCP) and minimum spanning tree approaches (MST) proposed by Lochhead et al [-@lochhead_demo_2018; -@lochhead_linking_2022] for solving the MTAP to another iterative least cost path method developed by Hardy [-@hardy_forest_2019; -@hardy_landis-ii_2023] (Figure \@ref(fig:proj-meth-fig)). We use the QGIS version of Hardy’s method (2019), which is easier for us to apply and does not differ substantially from the LANDIS extension (Hardy personal communication).

The ILCP approaches solve the MTAP by decomposing the problem into several smaller single target access problems (STAPs). For each landing target, the least cost path to the existing road network is solved using Dijkstra's shortest path algorithm [-@dijkstra_note_1959]. There are a variety of possible methods for building graphs, choosing branch locations, and ordering targets, and different methods will produce different networks [@anderson_projecting_2004; @hardy_forest_2019; @hardy_landis-ii_2023; @picard_finding_2006]. The ILCP algorithm we use [@lochhead_demo_2018; @lochhead_linking_2022] is parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction. Each node is connected with up to eight neighbours, and the cost of construction along edges is either the average of the cost raster value at the source and destination nodes, or a more complex function of the difference in elevation between the source and destination nodes (Table \@ref(tab:cost-tbl), Supplement A), weighted the distance to neighbouring nodes. Both the grade penalty and simple cost methods involve setting edge weights on a simple regular isotropic graph from adjacent locations on a single input grid; additional consideration of curvature and switchback spacing [@stuckelberger_improved_2007; @anderson_projecting_2004] would require additional computation. We use a greedy heuristic method that starts with the landing nearest to an existing road, builds a least cost path from the landing to the nearest point on the existing road, and updates the existing road network and weight raster with each new road segment before moving on to the next nearest landing.

Hardy’s method [-@hardy_forest_2019; -@hardy_landis-ii_2023] is similar to our simple cost ILCP method in that it breaks the MTAP down into many STAPs and finds the least cost path for each with the ordering of road construction determined by the Euclidean distance from the existing road network. A key difference is that Hardy uses the Dijkstra algorithm to solve a “single-source many-targets shortest path problem”, which is a special case of STAPs that finds the path between the harvested cell and one of multiple potential connection points on the road network. In contrast, our algorithm first finds the closest point on the road network to the selected landing point and then uses the Dijkstra algorithm to solve least cost path between the two. However, since the cost of construction on existing roads is zero we expect that in practice these two approaches are similar and the algorithm will select the least cost location for the road to connect to the existing road network. Within cutblocks, Hardy et al's method iterates over each cell in the harvest area and builds a road if it is not within skidding distance of an existing road, while we create landing points to serve as target locations for road construction.

Minimum spanning trees (MST) can also be used to link landings and evaluate road branching opportunities [@picard_finding_2006; @clark_three-stage_2000]. MST algorithms do not require users to select the order in which landings are processed, but it remains necessary to select a set of points on the existing road network to include in the tree. The first step in our MST approach is to estimate the least cost paths that link each of the landings and the least cost paths that link each landing to the nearest point on the existing road network. These results are used to create a weighted graph wherein nodes are either landings or nearest points on the existing road network, and edge weights correspond to the total cost of the least cost path between pairs of nodes. The minimum spanning tree for this graph is solved using Kruskal's algorithm [-@kruskal_shortest_1956], which links all nodes with a tree that minimizes edge weights. This method differs from Clark et al. [-@clark_three-stage_2000] who used Euclidean distances for edge weights, and thus did not incorporate spatial variation in road construction costs.

@@

To help make these road network projection methods more easily and widely useable, we implemented the ILCP and MST methods in the open source R package `roads` [@endicott_roads_2023]. The `roads` package is dependent on several other R packages including: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @csardi_igraph_2024]; `terra` for manipulating spatial raster data [@hijmans_terra_2023]; and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022]. We experimented with using graph construction methods in the package `gdistance` [@van_etten_r_2017], but opted to use Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] more computationally efficient methods (Supplement E) by default. The package is available on CRAN (The Comprehensive R Archive Network) and the source code, tutorials and documentation are available on GitHub [@endicott_roads_2023].

@@

"
df2 <- read.delim(textConnection(tbl2),header=FALSE,sep="|",strip.white=TRUE,stringsAsFactors=FALSE)
names(df2) <- unname(as.list(df2[1,]))
df2 <- df2[-1,] # remove first row
row.names(df2)<-NULL
knitr::kable(df2, caption = "Road projection method variants. We examined sensitivity of minimimum spanning tree (MST) and iterative least cost path (ILCP) algorithms to the method of setting edge weights (grade penalty or simple cost), the placement of landings within cutblocks (regular or random), the density of landings within cutblocks, and map resolution.")
```

@@

Although some perform better than others, all these methods underproject road density, road presence and forestry disturbance footprint (Figure \@ref(fig:aspat-perf-fig)). In other words, these road network projection methods project fewer roads than we observe. Projections of distance to road, the forestry disturbance footprint, and the road disturbance footprint are more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig)), suggesting that these measures are easier to project. One reason for underprojection of road density is that all these methods generate straight road segments within pixels. As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the weight raster resolution is coarse (Figure D.1).

@@

These road projection methods require construction of large graphs, with a node for each pixel in which road construction is possible. The associated memory and computational requirements may present challenges on large landscapes (Supplement E). Options for reducing computational requirements include restricting the area where road construction is possible (by setting some portions of the weight raster to NA), reducing landing density within cutblocks (by using the centroid method), increasing raster resolution, or using Hardy et al.'s more efficient implementations [-@hardy_forest_2019; -@hardy_landis-ii_2023]. The first three options are available to users of the `roads` R package. However, those contemplating reducing computational cost by increasing raster resolution should keep in mind that within grid cells all of these methods project straight roads, so in complex landscapes projections of road density become less accurate as raster cell resolution increases (Supplement D). The distance to road metric is less sensitive to changes in cell resolution, and distance metrics may also be more ecologically meaningful [@rowland_elk_2000]. Those considering reducing computational cost by reducing landing density within cutblocks should be clear that their projected road network will be incomplete even if it is also adequate for projecting the combined footprint of roads and cutblocks. More labour intensive possibilities for reducing computational costs include implementing slow components of the R package (Supplement F) in a faster compiled language, and implementing a multi-scale approach in which coarse resolution results are used to limit the area considered in finer resolution projections.

There are many metrics of connectivity and fragmentation [@francis_routledge_2021] we did not consider in this paper. In general, performance will vary among metrics, and it is difficult to determine what is good enough without reference to particular circumstances (i.e. objectives, species, landscapes). The most informative performance metrics will be those most closely linked to outcomes of interest. For example, where the outcome of interest is space use or connectivity projected from more [e.g. @whittington_towns_2022; @labadie_global_2023] or less [e.g. @diniz_landscape_2020; @hughes_comparison_2023] species-specific movement models it may be helpful to test sensitivity of these models to the difference between observed and retrospectively projected roads. When constructing new wildlife response models intended for projection, it may also be helpful to select metrics such as distance to road or buffered disturbance footprint that are more easily projectable [@bodner_bridging_2021]. Rather than adding additional performance metrics to this paper, we have focused our efforts on development of an open source R package [@endicott_roads_2023] and workflows [@endicott_roadpaper_2024] that we hope will help enable users to devise performance tests appropriate to their particular circumstances and goals. We recognize that acquiring data to retrospectively project roads can be challenging, so we also provide output maps [@endicott_assessment_2024] and an example script (analysis/scripts/4_calc_any_metric.R) to enable calculation of alternative performance metrics of interest for our Revelstoke example, along with a data preparation script (analysis/scripts/1_prepareData.R) that could be modified to extract required input data for other areas in British Columbia.

>>>>

institute:
  - ECCC: Landscape Science and Technology Division, National Wildlife Research Center, Environment and Climate Change Canada, Ottawa, ON, Canada  
  - FLNR: Ministry of Forest Lands, Natural Resource Operations and Rural Development, Victoria, BC, Canada
  - UBC:  Faculty of Forestry, Department of Forest Resources Management, The University of British Columbia, Vancouver, BC, Canada

@@

We expect these simple graph-based road projection methods to represent a reasonable compromise between computational cost and realism for stochastic projections on large landscapes, and we expect our simplified method for penalizing steep grades to yield more realistic road networks in steep terrain. We do not have a similarly strong hypothesis about whether minimum spanning tree projections are likely to be more or less realistic than iterative least cost path projections. If designers of real roads take a larger view to accessing multiple sites at the lowest cost we might expect minimum spanning tree algorithms to yield more realistic networks. However, real road networks are built over time, and it is not necessarily true that they are designed to minimize overall cost, so it is possible that larger more costly road networks built by iterative least-cost path algorithms will be more realistic.

@@

We chose the Revelstoke timber supply area (TSA 27) in British Columbia, Canada as an assessment area because it is mountainous (Figure \@ref(fig:study-area)), containing approximately 38% non-forested land [@british_columbia_ministry_of_forests_revelstoke_2022] where road building is difficult or impossible. Thus, the area provides opportunities to observe the behaviour of road projection algorithms in steep complex terrain. Forestry is the dominant resource extraction activity and much of the road network was built to access forest harvest cutblocks, allowing projection of the road network from historic forest harvest data. The Revelstoke TSA includes portions of the Monashee and Selkirk mountain ranges, and is bisected by the Columbia River. At low to mid-elevations, forests of the interior cedar-hemlock biogeoclimatic zone [@british_columbia_ministry_of_forests_revelstoke_2022] include remnants of rare old growth inland temperate rainforest [@government_of_canada_parks_canada_agency_old_2017]. At higher elevations, forests are dominated by Engelmann spruce and subalpine fir [@british_columbia_ministry_of_forests_revelstoke_2022]. The area is within the ancestral territories of the Ktunaxa Nation, Secwepemc Nation, Syilx and Sinixt [@british_columbia_ministry_of_forests_revelstoke_2022]. Two of six southern mountain woodland caribou herds in the area have been extirpated, and the remaining herds are considered Endangered by the Committee on the Status of Endangered Wildlife in Canada and listed as threatened under the Canadian Species at Risk Act [@palm_long_2020; @british_columbia_ministry_of_forests_revelstoke_2022; @environment_canada_recovery_2014]. Mountain caribou avoid roads leading to a loss of available habitat [@polfus_identifying_2011; @apps_factors_2006], suffer increased mortality near roads [@apps_spatial_2013], and decline in abundance as the footprint of roads and cutblocks increases [@lochhead_linking_2022]. Roads also alter grizzly bear habitat use, home range selection, movements, population fragmentation, survival, reproductive rates, population density, trends, and conservation status [@proctor_effects_2020].

@@

We focus on methods that link multiple target locations to an existing road network by finding minimum-cost locations for new roadways that link landings within harvest cutblocks (i.e., log decks) to the existing road network, described by Dean [-@dean_finding_1997] as the multiple target access problem (MTAP). Specifically, we compare iterative least cost path (ILCP) and minimum spanning tree approaches (MST) proposed by Lochhead et al [-@lochhead_demo_2018; -@lochhead_linking_2022] for solving the MTAP to another iterative least cost path method developed by Hardy [-@hardy_forest_2019; -@hardy_landis-ii_2023] (Figure \@ref(fig:proj-meth-fig)). We use the QGIS version of Hardy’s method (2019), which is easier for us to apply and does not differ substantially from the LANDIS extension (Hardy, personal communication).

The ILCP approaches solve the MTAP by decomposing the problem into several smaller single target access problems (STAPs). For each landing target, the least cost path to the existing road network is solved using Dijkstra's shortest path algorithm [@dijkstra_note_1959]. There are a variety of possible methods for building graphs, choosing branch locations, and ordering targets, and different methods will produce different networks [@anderson_projecting_2004; @hardy_forest_2019; @hardy_landis-ii_2023; @picard_finding_2006]. The ILCP algorithm we use [@lochhead_demo_2018; @lochhead_linking_2022] is parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction. Each node is connected with up to eight neighbours, and the cost of construction along edges is either the average of the cost raster value at the source and destination nodes, or a more complex function of the difference in elevation between the source and destination nodes (Table \@ref(tab:cost-tbl), Supplement A), weighted the distance to neighbouring nodes. Both the grade penalty and simple cost methods involve setting edge weights on a simple regular isotropic graph from adjacent locations on a single input grid; additional consideration of curvature and switchback spacing [@stuckelberger_improved_2007; @anderson_projecting_2004] would require additional computation. We use a greedy heuristic method that starts with the landing nearest to an existing road, builds a least cost path from the landing to the nearest point on the existing road, and updates the existing road network and weight raster with each new road segment before moving on to the next nearest landing.

The method described in Hardy et al. [-@hardy_forest_2019; -@hardy_landis-ii_2023] is similar to our simple cost ILCP method in that it breaks the MTAP down into many STAPs and finds the least cost path for each with the ordering of road construction determined by the Euclidean distance from the existing road network. A key difference is that Hardy uses the Dijkstra algorithm to solve a “single-source many-targets shortest path problem”, which is a special case of STAPs that finds the path between the harvested cell and one of multiple potential connection points on the road network. In contrast, our algorithm first finds the closest point on the road network to the selected landing point and then uses the Dijkstra algorithm to solve the least cost path between the two. However, since the cost of construction on existing roads is zero we expect that in practice these two approaches are similar and the algorithm will select the least cost location for the road to connect to the existing road network. Within cutblocks, the Hardy et al. method iterates over each cell in the harvest area and builds a road if it is not within skidding distance of an existing road, while we create landing points to serve as target locations for road construction.

Minimum spanning trees (MST) can also be used to link landings and evaluate road branching opportunities [@picard_finding_2006; @clark_three-stage_2000]. MST algorithms do not require users to select the order in which landings are processed, but it remains necessary to select a set of points on the existing road network to include in the tree. The first step in our MST approach is to estimate the least cost paths that link each of the landings and the least cost paths that link each landing to the nearest point on the existing road network. These results are used to create a weighted graph where nodes are either landings or nearest points on the existing road network, and edge weights correspond to the total cost of the least cost path between pairs of nodes. The minimum spanning tree for this graph is solved using Kruskal's algorithm [@kruskal_shortest_1956], which links all nodes with a tree that minimizes the sum of edge weights. This method differs from Clark et al. [-@clark_three-stage_2000] who used Euclidean distances for edge weights, and thus did not incorporate spatial variation in road construction costs.

@@

To help make these road network projection methods more easily and widely useable, we implemented the ILCP and MST methods in the open source R package `roads` [@endicott_roads_2023]. The `roads` package is dependent on several other R packages including: `igraph` for calculating least cost paths and minimum spanning trees [@csardi_igraph_2006; @csardi_igraph_2024]; `terra` for manipulating spatial raster data [@hijmans_terra_2023]; and `sf` for manipulating spatial vector data [@pebesma_simple_2018; @pebesma_sf_2022]. We experimented with using graph construction methods in the package `gdistance` [@van_etten_r_2017], but opted to use Lochhead et al's [-@lochhead_demo_2018; -@lochhead_linking_2022] more computationally efficient methods (Supplement F) by default. The package is available on CRAN (The Comprehensive R Archive Network) and the source code, tutorials and documentation are available on GitHub [@endicott_roads_2023].

@@

"
df2 <- read.delim(textConnection(tbl2),header=FALSE,sep="|",strip.white=TRUE,stringsAsFactors=FALSE)
names(df2) <- unname(as.list(df2[1,]))
df2 <- df2[-1,] # remove first row
row.names(df2)<-NULL
knitr::kable(df2, caption = "Road projection method variants. We examined sensitivity of minimum spanning tree (MST) and iterative least cost path (ILCP) algorithms to the method of setting edge weights (grade penalty or simple cost), the placement of landings within cutblocks (regular or random), the density of landings within cutblocks, and map resolution.")
```

@@

Although some perform better than others, all these methods underproject road density, road presence and forestry disturbance footprint (Figure \@ref(fig:aspat-perf-fig)). In other words, these road network projection methods project fewer roads than we observe. Projections of distance to road, the forestry disturbance footprint, and the road disturbance footprint are more accurate than projections of road density and road presence (Figure \@ref(fig:aspat-perf-fig)), suggesting that these measures are easier to project. One reason for underprojection of road density is that all these methods generate straight road segments within pixels. As raster resolution increases, the discrepancy within pixels between real roads and straight projected segments increases, which explains why projections of road density are less accurate when the raster resolution is coarse (Figure D.1).

@@

These road projection methods require construction of large graphs, with a node for each pixel in which road construction is possible. The associated memory and computational requirements may present challenges on large landscapes (Supplement E). Options for reducing computational requirements include restricting the area where road construction is possible (by setting some portions of the weight raster to NA), reducing landing density within cutblocks (by using the centroid method), increasing raster resolution, or using Hardy's more efficient implementations. The first three options are available to users of the `roads` R package. However, those contemplating reducing computational cost by decreasing raster resolution should keep in mind that within grid cells all of these methods project straight roads, so in complex landscapes projections of road density become less accurate as raster cell resolution increases (Supplement D). The distance to road metric is less sensitive to changes in cell resolution, and distance metrics may also be more ecologically meaningful [@rowland_elk_2000]. Those considering reducing computational cost by reducing landing density within cutblocks should be clear that their projected road network will be incomplete even if it is also adequate for projecting the combined footprint of roads and cutblocks. More labour intensive possibilities for reducing computational costs include implementing slow components of the R package (Supplement F) in a faster compiled language, and implementing a multi-scale approach in which coarse resolution results are used to limit the area considered in finer resolution projections.

There are many metrics of connectivity and fragmentation [@francis_routledge_2021] we did not consider in this paper. In general, performance will vary among metrics, and it is difficult to determine what is good enough without reference to particular circumstances (i.e. objectives, species, landscapes). The most informative performance metrics will be those most closely linked to outcomes of interest. For example, where the outcome of interest is space use or connectivity projected from more [e.g. @whittington_towns_2022; @labadie_global_2023] or less [e.g. @diniz_landscape_2020; @hughes_comparison_2023] species-specific movement models it may be helpful to test sensitivity of these models to the difference between observed and retrospectively projected roads. When constructing new wildlife response models intended for projection, it may also be helpful to select metrics such as distance to road or buffered disturbance footprint that are more easily projectable [@bodner_bridging_2021]. Rather than adding additional performance metrics to this paper, we have focused our efforts on development of an open source R package [@endicott_roads_2023] and workflows [@endicott_roadpaper_2024] that we hope will help enable users to devise performance tests appropriate to their particular circumstances and goals. We recognize that acquiring data to retrospectively project roads can be challenging, so we also provide output maps [@endicott_assessment_2024] and an example script (analysis/scripts/4_calc_any_metric.R) to enable calculation of alternative performance metrics of interest for our Revelstoke example, along with a data preparation script (analysis/scripts/1_prepareData.R) that could be modified to extract required input data for other areas in British Columbia [@endicott_roadpaper_2024].
